prolfqua - A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-06-03 inst/MyArticle/paper.Rmd
more edits to implementation.

We store all the data needed for analysis in a single data frame in a tidy table, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an __R6__[@R6cite] configuration object \@ref(fig:LFQData), we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy table. For example, to visualize tidy _Spectronaut_, _DiaNN_, or _Skyline_ outputs, or data in `r BiocStyle::Biocpkg("MSstats")` format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` configuration. The configuration encapsulates the differences of column names among the various input formats and enables the methods of prolfqua to be transparent, i.e., it preserves the differences of the processing tools in the figures and analysis. We show an example code for creating an MSFragger configuration in the Supplement. For popular software like _MaxQuant_, or_MSFragger_, which stores the same variable, e.g., intensity, in multiple columns, one for each sample, we implemented methods that transform the data into tidy tables. Furthermore, relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use __R6__ classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). __R6__ classes are well supported, e.g., by command-line completion features, see Figure \@ref(fig:tabCompletion), and easy to use (OOP).

@@

In addition we implemented features specific to high throughput experiments, such as the experimental Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We used R6 classes encapsulate the statistical modelling functionality in `r BiocStyle::Biocpkg("prolfqua")` (see Figure \@ref(fig:ContrastUML)). We did specify a contrast interface (ContrastInterface), and there are various implementations allowing to perform differential expression analysis given linear or mixed effect models (Contrast), variance shrinkage (ContrastModerated), or impute contrasts in case of observations missing for an entire group of samples (ContrastSimpleImpute) can also be used to encapsulate and integrate differential expression analysis results of external tools such as proDA or to analyse data from protein interaction studies using SAINTexpress [@teo2014saintexpress].

(ref:ContrastUML) UML diagram of modeling and contrast related classes. Different strategies, e.g., _lm_, _lmer_, and _glm_ can be used to fit various types of models. The model builder method fit's the statistical model given the data and a strategy. The obtained model can then be used for analysis of variance or to estimate contrasts. All classes estimating the contrasts implement the _Contrast_ interface. The variance of the contrasts can be moderated using the `ContrastModerated` class.

@@

colnames(xM) <- c("prolfqua functions", "model")
knitr::kable(tibble::as_tibble(xM), caption = "Prolfqua functions to fit various models.",
             format = "latex", booktabs=TRUE) |>
    kableExtra::kable_styling(latex_options="scale_down") |> 
    kableExtra::add_footnote(label = "* in development")

@@

xMM <- data.frame(t(xMM))
rownames(xMM) <- NULL
colnames(xMM) <-  c("Label","Description","Abundance","Input File")
kableExtra::kable(xMM,
             caption = paste0("All benchmarked models.",
             "Description - prolfqua function names",
             "Abudances - indicates if model is fitted to peptide or protein abundances",
             "Input File - MaxQuant file used as input to the analysis."),
             format = "latex", booktabs=TRUE) |>
    kableExtra::kable_styling(latex_options="scale_down") |> 
    kableExtra::add_footnote(label = "* Merges results of the prot_med_lm_moderated and prot_imputation modeling pipeline, preferring those of prot_med_lm_moderated if available.")

@@

The probabilistic dropout analysis implemented in the `r BiocStyle::Biocpkg("proDA")` produces inferences comparable to those of other methods (Figure \@ref(fig:FDRfdp) Panel A) for the largest number of proteins (Figure \@ref(fig:FDRfdp) Panel B). Moreover, the estimated differences are unbiased (not shown) and show high diagnostic accuracy (\@ref(fig:FDRfdp) Panel A). Furthermore, the performance of the scaled $p$-values or the $t$-statistics is comparable with that of the linear model with variance moderation. Because of the robustness of the dropout model to missing observations, we intend to integrate the `r BiocStyle::Biocpkg("proDA")` package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)).

@@

`r BiocStyle::Githubpkg("fgcz/prolfqua")` allows for considerable flexibility to model your experiments. Various types of models are available (see Figure \@ref(fig:ContrastUML) and Table \@ref(tab:prolfquaModels)), and the contrast specification is consistent and intuitive for all models. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")`, allows for adding new features, e.g., generalized linear models (_glm_'s) for missing data modeling or robust linear models (_rlm_'s), in the future. R's formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending]. We use the formula interface to specify the models which should make it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical programming languages.

>>>>

We store all the data needed for analysis in a tidy table, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an __R6__[@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy table. For example, to visualize tidy _Spectronaut_, _DiaNN_, or _Skyline_ outputs, or data in `r BiocStyle::Biocpkg("MSstats")` format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences in column names among the various input formats, enables the use of the methods without additional arguments, and preserves the processing tools' differences in the figures and analysis. We show an example code for creating an MSFragger configuration in the Supplement. For popular software like _MaxQuant_, or_MSFragger_, which stores the same variable, e.g., intensity, in multiple columns, one for each sample, we implemented methods that transform the data into tidy tables. Furthermore, relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use __R6__ classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). __R6__ classes are well supported, e.g., by command-line completion features, see Figure \@ref(fig:tabCompletion), and easy to use (OOP).

@@

In addition, we implemented features specific to high throughput experiments, such as the experimental Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We used R6 classes to encapsulate the statistical modelling functionality in `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)). We did specify a contrast interface (`ContrastsInterface`). Several implementations allow to perform differential expression analysis given linear or mixed effect models (`Contrasts`), variance shrinkage (`ContrastsModerated`), or impute contrasts in cases when observations are missing for an entire group of samples (`ContrastsSimpleImpute`). Further implementations of the interface encapsulate and integrate differential expression analysis results of external tools such as `r `BiocStyle::Biocpkg("proDA")` or of SAINTexpress [@teo2014saintexpress] used to analyze data from protein interaction studies using.

(ref:ContrastUML)  UML diagram of modeling and contrast related classes. Different strategies, e.g., _lm_, _lmer_, and _glm_, can fit various types of models. The model builder method fit's the statistical model given the data and a strategy. The obtained model can then be used for the analysis of variance or to estimate contrasts. All classes estimating contrasts implement the _Contrasts_ interface. Results of external tools, e.g., SAINTExpress, can be adapted by implementing the Contrasts Interface.

@@

colnames(xM) <- c("prolfqua functions", "model")
knitr::kable(tibble::as_tibble(xM), caption = "Prolfqua functions to fit various models.",
             format = "latex", booktabs=TRUE) |>
    kableExtra::kable_styling(latex_options="scale_down") |> 
    kableExtra::add_footnote(label = "in development")

@@

xMM <- data.frame(t(xMM))
rownames(xMM) <- NULL
colnames(xMM) <-  c("Label","Description","Abundance","Input File")
kableExtra::kable(xMM,
             caption = paste0("All benchmarked models.",
             "Description - prolfqua function names",
             "Abudances - indicates if model is fitted to peptide or protein abundances",
             "Input File - MaxQuant file used as input to the analysis."),
             format = "latex", booktabs=TRUE) |>
    kableExtra::kable_styling(latex_options="scale_down") |> 
    kableExtra::add_footnote(label = "Merges results of the prot_med_lm_moderated` and prot_imputation modeling pipeline, preferring those of prot_med_lm_moderated if available.")

@@

The probabilistic dropout analysis implemented in the `r BiocStyle::Biocpkg("proDA")` produces inferences comparable to those of other methods (Figure \@ref(fig:FDRfdp) Panel A). Because of the robustness of the dropout model to missing observations we obtain difference estimates for all proteins and contrasts (Figure \@ref(fig:FDRfdp) Panel B). Moreover, the estimated differences are unbiased and show high diagnostic accuracy (\@ref(fig:FDRfdp) Panel A). Furthermore, the performance of the scaled $p$-values or the $t$-statistics is comparable with that of the linear model with variance moderation (`prot_med_lm_moderated`). Therefore, we started to integrate the `r BiocStyle::Biocpkg("proDA")` package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)).

@@

`r BiocStyle::Githubpkg("fgcz/prolfqua")` allows for considerable flexibility to model your experiments. Various types of models are available (see Figure \@ref(fig:ContrastUML) and Table \@ref(tab:prolfquaModels)), and the contrast specification is explicit and consistent for all models. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")`, allows for adding new features, e.g., generalized linear models (_glm_'s) to model the presence or absence information of a protein, or robust linear models (_rlm_'s), in the future. R's formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending]. We use the formula interface to specify the models, making it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical methods programming languages. The developed framework is flexible to integrate other modeling methods, e.g., the probabilistic dropout model [@bioxrvproDA2020] or accurate variance estimation [@zhu2020deqms].
